---
title: "Assessing the effectiveness of two intervention methods for stony coral tissue loss disease on *Montastraea cavernosa*"
author: "Erin Shilling -- eshilling15@gmail.com, Ian Combs -- combsi9892@gmail.com, and Dr. Joshua D. Voss -- jvoss2@fau.edu "
date: "2/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
***
# About this document
***
All analyses were performed in R version 3.5.2. This is the code that accompanies the publication (Shilling EN, Combs IR, Voss JD (2021) Assessing the effectiveness of two intervention methods for stony coral tissue loss disease on *Montastraea cavernosa*. *Sci Rep*). Here you will find all the code to repeat the statistical analyses performed in R for the manuscript.

Naming conventions:

Treatment groups
"Chlorine" & "Chl" both refer to the chlorinated epoxy treatment. "Amoxicillin" & "Amox" refer to the Base 2B plus amoxicillin treatment. "Untreated" & "Untr" refer to the SCTLD-affected, untreated control group.

Time points
"TP0" refers to the initial experiment set up, with "Immediately before treatment" & "Pre" referring to the chlorine and amoxicillin treated colonies before treatment at experiment setup, and "Immediately after treatment" and "Post" referring to the chlorine and amoxicillin treated colonies right after treatment at experiment setup. Untreated controls were also included in this "Immediately after treatment" & "Post" subset.
"TP1" refers to the first follow-up event that occurred 3 weeks after initial interventions.
"TP2" refers to the second follow-up event that occurred 5 weeks after initial interventions.
"TP3" refers to the third follow-up event that occurred 9 weeks after initial interventions.
"TP4" refers to the fourth follow-up event that occurred 14 weeks after initial interventions.
"TP5" refers to the fifth follow-up event that occurred 23 weeks after initial interventions.
"TP6" refers to the sixth follow-up event that occurred 46 weeks after initial interventions.

If you have any issues or questions about the code please feel free to send an email to eshilling15@gmail.com.

```{r, loading required packages and datasets}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(Rmisc, psych, tidyverse, rstatix, FSA)

Intervention.Data <- read.csv("https://raw.githubusercontent.com/erin-shilling/SCTLDintervention_ScientificReports/main/Intervention.Data.csv")
```

```{r, subsetting data}

#R doesn't automatically recognize as numeric due to NAs
Intervention.Data$Initial.Colony.Area <- as.numeric(as.character(Intervention.Data$Initial.Colony.Area))
Intervention.Data$Initial.TP0.Lesions <- as.numeric(as.integer(Intervention.Data$Initial.TP0.Lesions))

#make dataframe excluding TP0
Intervention.Data.NoTP0 <- Intervention.Data[which(Intervention.Data$Time.Point.Num>0, ), ]

#make data frame without colony 161
Intervention.Data.No161 <- Intervention.Data[which(Intervention.Data$Colony.ID!=161, ), ]

#make data frame without colonies that died through course of experiment
Intervention.Data.No.Dead <- Intervention.Data.NoTP0[which(Intervention.Data.NoTP0$Current.Health.Status!= "Dead", ), ]

#subset by time point
TP1<-Intervention.Data[Intervention.Data$Time.Point=="First followup", ]
TP2<-Intervention.Data[Intervention.Data$Time.Point=="Second followup", ]
TP3<-Intervention.Data[Intervention.Data$Time.Point=="Third followup", ]
TP4<-Intervention.Data[Intervention.Data$Time.Point=="Fourth followup", ]
TP5<-Intervention.Data[Intervention.Data$Time.Point=="Fifth followup", ]
TP6<-Intervention.Data[Intervention.Data$Time.Point=="Sixth followup", ]

#no dead colonies included
TP6.No.Dead<-Intervention.Data.No.Dead[Intervention.Data.No.Dead$Time.Point=="Sixth followup", ]

#subset dataframe without TP0 by treatment
Amox.NoTP0 <- Intervention.Data.NoTP0[Intervention.Data.NoTP0$Treatment=="Amoxicillin", ]
Chl.NoTP0 <- Intervention.Data.NoTP0[Intervention.Data.NoTP0$Treatment=="Chlorine", ]
Untr.NoTP0 <- Intervention.Data.NoTP0[Intervention.Data.NoTP0$Treatment=="Untreated", ]

```

```{r, summarizing data sets}

#initial surface area data
describe(Intervention.Data$Initial.Colony.Area)

```

```{r, assessing normality}
#initial colony size data
##time point used doesn't matter since it's just initial colony area

shapiro.test(TP6$Initial.Colony.Area)
TP6 %>% group_by(Treatment) %>% shapiro_test(Initial.Colony.Area)
#only one treatment of the 3 is normal, transformations unsuccessful

#total lesions at 46 weeks by site
TP6 %>% group_by(Site) %>% shapiro_test(Total.current.lesions)
#none are normal

#cumulative new lesions at 46 weeks by site
TP6 %>% group_by(Site) %>% shapiro_test(Cumulative.New.Lesions)
#none are normal

#New lesions developed at each time point, by treatment
##For time points 3 and 6, some treatment groups had no new lesions at all, so a shapiro test by group was not possible
TP1 %>% group_by(Treatment) %>% shapiro_test(New.Lesions)
TP2 %>% group_by(Treatment) %>% shapiro_test(New.Lesions)
shapiro_test(TP3$New.Lesions)
TP4 %>% group_by(Treatment) %>% shapiro_test(New.Lesions)
TP5 %>% group_by(Treatment) %>% shapiro_test(New.Lesions)
shapiro.test(TP6$New.Lesions)
```

```{r, Kruskal-Wallis: initial colony surface area differences across treatment groups}

kruskal.test(TP6$Initial.Colony.Area~TP6$Treatment)

```
No significant difference in mean initial colony area across treatment groups at start of experiment

```{r, Kruskal-Wallis: initial colony surface area differences across sites}

kruskal.test(TP6$Initial.Colony.Area~TP6$Site)
```
No significant difference in mean initial colony area across sites at start of experiment

```{r, Kruskal-Wallis: initial colony surface influence on TP6 SCTLD status of colonies}

kruskal.test(TP6$Initial.Colony.Area~TP6$Current.Health.Status)

##by treatment group

TP6 %>% group_by(Treatment) %>% kruskal_test(Initial.Colony.Area~Current.Health.Status)
```
No significant difference in mean initial colony areas of colonies when broken up by SCTLD status at TP6

```{r, Kruskal-Wallis: site influence on total lesions at 46 weeks}

kruskal.test(TP6.No.Dead$Total.current.lesions~TP6.No.Dead$Site)
```

No significant difference between mean total number of lesions present on colonies across sites

```{r, Kruskal-Wallis: site influence on cumulative new lesions at 46 weeks}

kruskal.test(TP6.No.Dead$Cumulative.New.Lesions~TP6.No.Dead$Site)
```
No significant difference between mean cumulative new lesions developed on colonies by 46 weeks across sites

```{r, Kruskal-Wallis: treatment influence on new lesions developed between each time point}

Intervention.Data.No.Dead %>% group_by(Time.Point.Num) %>% kruskal_test(New.Lesions~Treatment)

```
No influence of treatment on development of new lesions at any time point

```{r, Kruskal-Wallis: treatment influence on cumulative new lesions at 46 weeks}

kruskal.test(TP6.No.Dead$Cumulative.New.Lesions~TP6.No.Dead$Treatment)
```
No impact of treatment on cumulative new lesions developed by 46 weeks

```{r, Kruskal-Wallis: differences in number of initial lesions on colonies across sites}

kruskal.test(TP1$Initial.TP0.Lesions~TP1$Site)
```
No difference in mean number of initial lesions present on colonies across sites

```{r, Kruskal-Wallis: differences in number of initial lesions on colonies across treatment groups}

kruskal.test(TP1$Initial.TP0.Lesions~TP1$Treatment)
```

No difference in mean number of initial lesions present on colonies across treatment groups

```{r, Fisher Test: site influence on SCTLD status of colonies at any time point}

#TP1
##get counts
TP1 %>% count(Site, Current.Health.Status)
##make table
TP1.status.site <- as.table(rbind(c(9, 4, 9), c(7, 0, 3)))
dimnames(TP1.status.site) <- list(Quiesced = c("No", "Yes"),Site = c("BC1", "FTL4", "T328"))

##run test
fisher_test(TP1.status.site)

#TP2
TP2 %>% count(Site, Current.Health.Status)
TP2.status.site <- as.table(rbind(c(9, 4, 10), c(7, 0, 2)))
dimnames(TP2.status.site) <- list(Quiesced = c("No", "Yes"),Site = c("BC1", "FTL4", "T328"))

fisher_test(TP2.status.site)

#TP3
TP3 %>% count(Site, Current.Health.Status)
TP3.status.site <- as.table(rbind(c(10, 4, 11), c(6, 0, 1)))
dimnames(TP3.status.site) <- list(Quiesced = c("No", "Yes"),Site = c("BC1", "FTL4", "T328"))

fisher_test(TP3.status.site)

#TP4
TP4 %>% count(Site, Current.Health.Status)
TP4.status.site <- as.table(rbind(c(9, 4, 11), c(7, 0, 1)))
dimnames(TP4.status.site) <- list(Quiesced = c("No", "Yes"),Site = c("BC1", "FTL4", "T328"))

fisher_test(TP4.status.site)

#TP5
TP5 %>% count(Site, Current.Health.Status)
TP5.status.site <- as.table(rbind(c(10, 3, 11), c(6, 1, 1)))
dimnames(TP5.status.site) <- list(Quiesced = c("No", "Yes"),Site = c("BC1", "FTL4", "T328"))

fisher_test(TP5.status.site)

#TP6
TP6 %>% count(Site, Current.Health.Status)
TP6.status.site <- as.table(rbind(c(6, 3, 9), c(10, 1, 3)))
dimnames(TP6.status.site) <- list(Quiesced = c("No", "Yes"),Site = c("BC1", "FTL4", "T328"))

fisher_test(TP6.status.site)
```
No impact of site on status of colonies at any time point

```{r, Fisher Test: influence of treatment on colony SCTLD status at each time point}

#TP1
TP1 %>% count(Treatment, Current.Health.Status)

TP1.status.treat <- as.table(rbind(c(2, 10, 10), c(9, 1, 0)))
dimnames(TP1.status.treat) <- list(Quiesced = c("No", "Yes"),Treatment = c("Amox", "Chlor", "Untr"))

fisher_test(TP1.status.treat)

pairwise_fisher_test(TP1.status.treat, p.adjust.method = "bonferroni")

#TP2
TP2 %>% count(Treatment, Current.Health.Status)

TP2.status.treat <- as.table(rbind(c(3, 10, 10), c(8, 1, 0)))
dimnames(TP2.status.treat) <- list(Quiesced = c("No", "Yes"),Treatment = c("Amox", "Chlor", "Untr"))

fisher_test(TP2.status.treat)

pairwise_fisher_test(TP2.status.treat, p.adjust.method = "bonferroni")

#TP3
TP3 %>% count(Treatment, Current.Health.Status)

TP3.status.treat <- as.table(rbind(c(4, 11, 10), c(7, 0, 0)))
dimnames(TP3.status.treat) <- list(Quiesced = c("No", "Yes"),Treatment = c("Amox", "Chlor", "Untr"))

fisher_test(TP3.status.treat)

pairwise_fisher_test(TP3.status.treat, p.adjust.method = "bonferroni")

#TP4
TP4 %>% count(Treatment, Current.Health.Status)

TP4.status.treat <- as.table(rbind(c(5, 11, 8), c(6, 0, 2)))
dimnames(TP4.status.treat) <- list(Quiesced = c("No", "Yes"),Treatment = c("Amox", "Chlor", "Untr"))

fisher_test(TP4.status.treat)

pairwise_fisher_test(TP4.status.treat, p.adjust.method = "bonferroni")

#TP5
TP5 %>% count(Treatment, Current.Health.Status)

TP5.status.treat <- as.table(rbind(c(6, 11, 7), c(5, 0, 3)))
dimnames(TP5.status.treat) <- list(Quiesced = c("No", "Yes"),Treatment = c("Amox", "Chlor", "Untr"))

fisher_test(TP5.status.treat)

pairwise_fisher_test(TP5.status.treat, p.adjust.method = "bonferroni")

#TP6
TP6 %>% count(Treatment, Current.Health.Status)

TP6.status.treat <- as.table(rbind(c(4, 8, 6), c(7, 3, 4)))
dimnames(TP6.status.treat) <- list(Quiesced = c("No", "Yes"),Treatment = c("Amox", "Chlor", "Untr"))

fisher_test(TP6.status.treat)
```
Treatment influenced colony SCTLD status significantly until the fifth follow-up event at 23 weeks. Until the third follow-up at 9 weeks, amoxicillin-treated colonies were more likely to be quiesced than chlorine-treated and untreated, by the fourth follow-up at 14 weeks amoxicillin-treated were more likely to be quiesced only when compared to chlorine-treated, and by 23 weeks, despite a significant global test, no pairwise comparisons between treatment groups were significant.

```{r, Fisher Test: influence of treatment on initially treated SCTLD lesion status at 46 weeks}

TP6 %>% group_by(Treatment) %>% tally(Initial.TP0.Lesions)
TP6 %>% group_by(Treatment) %>% tally(TP0.Lesions.active)

#making table for 3x2 Fisher test
treated.lesions <- as.table(rbind(c(34, 1, 18), c(7, 21, 14)))
dimnames(treated.lesions) <- list(Lesion.Quiesced = c("No", "Yes"),Treatment = c("Chlor", "Amox", "Untr"))

#run test
fisher_test(treated.lesions)

#pairwise comparisons
pairwise_fisher_test(treated.lesions, p.adjust.method = "bonferroni")
```
Significant impact of treatment on initially treated lesion status at 46 weeks
Amoxicillin-treated more likely to be quiesced than chlorine and untreated, chlorine vs. untreated no difference

```{r, Kruskal-Wallis: influence of initial number of lesions present on a colony SCTLD status at any time point}
##NOTE: I saw chunk options ", Kruskal-Wallis: influence of initial number lesions present on a colony's SCTLD status at any time point" please go to https://yihui.org/knitr/options (it is likely that you forgot to quote "character" options)

Intervention.Data.NoTP0 %>% group_by(Time.Point.Num) %>% kruskal_test(Current.Health.Status~Initial.TP0.Lesions)

Amox.NoTP0 %>% group_by(Time.Point.Num) %>% kruskal_test(Current.Health.Status~Initial.TP0.Lesions)

Chl.NoTP0 %>% group_by(Time.Point.Num) %>% kruskal_test(Current.Health.Status~Initial.TP0.Lesions)

Untr.NoTP0 %>% group_by(Time.Point.Num) %>% kruskal_test(Current.Health.Status~Initial.TP0.Lesions)

```
No impact of initial lesions present on a colony on it's SCTLD status at any time point

```{r, Fisher Test: difference in proportions of amoxicillin and chlorine treated colonies with tissue left between their trenched areas and initial disease margins at 46 weeks}

#make table for 2x2 Fisher test
trench.treat <- as.table(rbind(c(4, 17), c(10, 1)))
dimnames(trench.treat) <- list(Tissue.Left = c("No", "Yes"),Treatment = c("Amox", "Chlor"))

#run test
fisher_test(trench.treat)
```
Amoxicillin-treated colonies more likely than chlorine-treated to have tissue remaining at 46 weeks between trenched area and initial disease margins

```{r, Spearmans Rank Correlation, associations between time and the development of new lesions on a colony}

#look at Spearman's correlation between time and new lesions excluding TP0
##all treatments combined
scatter.smooth(x=Intervention.Data.NoTP0$Days.Since.Intervention, y=Intervention.Data.NoTP0$New.Lesions, main="New lesions ~ Time")
cor.test(Intervention.Data.NoTP0$New.Lesions, Intervention.Data.NoTP0$Days.Since.Intervention, method = "spearman")

#amox only
scatter.smooth(x=Amox.NoTP0$Days.Since.Intervention, y=Amox.NoTP0$New.Lesions, main="New lesions ~ Time")
#chlorine only
scatter.smooth(x=Chl.NoTP0$Days.Since.Intervention, y=Chl.NoTP0$New.Lesions, main="New lesions ~ Time")
#untreated only
scatter.smooth(x=Untr.NoTP0$Days.Since.Intervention, y=Untr.NoTP0$New.Lesions, main="New lesions ~ Time")

#Spearman's correlations again by treatment groups
Intervention.Data.NoTP0 %>% group_by(Treatment) %>% cor_test(New.Lesions, Days.Since.Intervention, method = "spearman")
```
No correlation between time and new lesions developed, even when blocked by treatment group

```{r, Spearmans Rank Correlation, associations between time and total lesions present on a colony}

#look at Spearman's correlation between time and total lesions excluding TP0
##all treatments combined
scatter.smooth(x=Intervention.Data.NoTP0$Days.Since.Intervention, y=Intervention.Data.NoTP0$Total.current.lesions, main="Total lesions ~ Time")
cor.test(Intervention.Data.NoTP0$Total.current.lesions, Intervention.Data.NoTP0$Days.Since.Intervention, method = "spearman")

#amox only
scatter.smooth(x=Amox.NoTP0$Days.Since.Intervention, y=Amox.NoTP0$Total.current.lesions, main="Total lesions ~ Time")
#chlorine only
scatter.smooth(x=Chl.NoTP0$Days.Since.Intervention, y=Chl.NoTP0$Total.current.lesions, main="Total lesions ~ Time")
#untreated only
scatter.smooth(x=Untr.NoTP0$Days.Since.Intervention, y=Untr.NoTP0$Total.current.lesions, main="Total lesions ~ Time")

#Spearman's correlations again by treatment groups
Intervention.Data.NoTP0 %>% group_by(Treatment) %>% cor_test(Total.current.lesions, Days.Since.Intervention, method = "spearman")
```
No correlation between time and total lesions present, even when blocked by treatment group

```{r, Spearmans Rank Correlation, associations between initial colony surface area and the initial number of SCTLD lesions present}

#correlation
scatter.smooth(x=Intervention.Data$Initial.Colony.Area, y=Intervention.Data$Initial.TP0.Lesions, main="initial lesions ~ size")
cor.test(Intervention.Data$Initial.Colony.Area, Intervention.Data$Initial.TP0.Lesions, method = "spearman")

#correlation without Colony 161
scatter.smooth(x=Intervention.Data.No161$Initial.Colony.Area, y=Intervention.Data.No161$Initial.TP0.Lesions, main="initial lesions ~ surface area")
cor.test(Intervention.Data.No161$Initial.Colony.Area, Intervention.Data.No161$Initial.TP0.Lesions, method = "spearman")

```
Significant positive correlation, slightly skewed by one outlier colony but still signficiant with outlier removed

```{r, Spearmans Rank Correlation: associations between initial colony surface area and new lesions developed}

scatter.smooth(x=TP6$Initial.Colony.Area, y=TP6$Cumulative.New.Lesions, main="Initial Area ~ New Lesions")
cor.test(TP6$Initial.Colony.Area, TP6$Cumulative.New.Lesions, method = "spearman")

```
